---
title: "Cyclistic Bike-share data analysis"
author: "Atien"
output: html_document
date: "2023-07-28"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Get Started

Loading packages needed in this analysis

```{r load packages}
library(tidyverse)
library(janitor)
library(skimr)
library(lubridate)
```

## Import Data

```{r Import and Combine Data}
data_202207 <- read.csv("202207-divvy-tripdata.csv")
data_202208 <- read_csv("202208-divvy-tripdata.csv")
data_202209 <- read_csv("202209-divvy-publictripdata.csv")
data_202210 <- read_csv("202210-divvy-tripdata.csv")
data_202211 <- read_csv("202211-divvy-tripdata.csv")
data_202212 <- read_csv("202212-divvy-tripdata.csv")
data_202301 <- read_csv("202301-divvy-tripdata.csv")
data_202302 <- read_csv("202302-divvy-tripdata.csv")
data_202303 <- read_csv("202303-divvy-tripdata.csv")
data_202304 <- read_csv("202304-divvy-tripdata.csv")
data_202305 <- read_csv("202305-divvy-tripdata.csv")
data_202306 <- read_csv("202306-divvy-tripdata.csv")
data_202306 <- read_csv('202306-divvy-tripdata.csv')

```

## Check Column Names
check whether there are different column names
```{r Check Column Names}
colnames(data_202207)
colnames(data_202208)
colnames(data_202209)
colnames(data_202210)
colnames(data_202211)
colnames(data_202212)
colnames(data_202301)
colnames(data_202302)
colnames(data_202303)
colnames(data_202304)
colnames(data_202305)
colnames(data_202306)
```

## Data United and Combined
combined all data into one dataframe
```{r Data United and Combined}
compare_df_cols(data_202207,data_202208,data_202209,data_202210,data_202211,data_202212,data_202301,data_202302,data_202303,data_202304,data_202305,data_202306,return = "mismatch")
data_202207$started_at <- as.POSIXct(data_202207$started_at,format="%Y-%m-%d %H:%M:%S")
data_202207$ended_at <- as.POSIXct(data_202207$ended_at,format="%Y-%m-%d %H:%M:%S")

data_202207_to_202306 <- bind_rows(data_202207,data_202208,data_202209,data_202210,data_202211,data_202212,data_202301,data_202302,data_202303,data_202304,data_202305,data_202306)
```

## Add Columns
add columns for later analysis
```{r Add Columns}
data_202207_to_202306$day_of_week <- wday(data_202207_to_202306$started_at,week_start=1)
data_202207_to_202306$month <- month(data_202207_to_202306$started_at)
data_202207_to_202306$trip_time <- difftime(data_202207_to_202306$ended_at, data_202207_to_202306$started_at, units="min")
data_202207_to_202306$trip_time <- round(data_202207_to_202306$trip_time,2)
data_202207_to_202306$trip_time <- as.numeric(as.character(data_202207_to_202306$trip_time))

```

## Quick Preview Table
```{r Quick Preview Table}
skim_without_charts(data_202207_to_202306)
```

## Data Clean
drop null value and filter data
```{r Data Clean}
data_cleaned <-data_202207_to_202306 %>%
  drop_na() %>%
  filter(start_station_name == "HQ QR" | trip_time>0) %>%
  filter(trip_time<1440 & trip_time>1) %>%
  filter(started_at < ended_at)

```
## Analyzing Data
Descriptive analysis
```{r Descriptive analysis}
summary(data_cleaned$trip_time)
```
Compare members and casual users
```{r Compare members and casual users}
aggregate(data_cleaned$trip_time ~ data_cleaned$member_casual, FUN = mean)
aggregate(data_cleaned$trip_time ~ data_cleaned$member_casual, FUN = median)
aggregate(data_cleaned$trip_time ~ data_cleaned$member_casual, FUN = max)
aggregate(data_cleaned$trip_time ~ data_cleaned$member_casual, FUN = min)

# See the average ride time by each day for members vs casual users
aggregate(data_cleaned$trip_time ~ data_cleaned$member_casual + data_cleaned$day_of_week, FUN = mean)

# The most commonly used start station for casual user
popular_start_casual <- data_cleaned %>%
  group_by(member_casual,start_station_name)%>%
  count(start_station_name)%>%
  arrange(desc(n)) %>%
  filter(member_casual=='casual')
head(popular_start_casual)

# The most commonly used start station for member
popular_start_member <- data_cleaned %>%
  group_by(member_casual,start_station_name)%>%
  count(start_station_name)%>%
  arrange(desc(n)) %>%
  filter(member_casual=='member')
head(popular_start_member)

# The most commonly used end station for casual user
popular_end_casual <- data_cleaned %>%
  group_by(member_casual,end_station_name)%>%
  count(end_station_name)%>%
  arrange(desc(n))%>%
  filter(member_casual=='casual')
head(popular_end_casual)

# The most commonly used end station for member
popular_end_member <- data_cleaned %>%
  group_by(member_casual,end_station_name)%>%
  count(end_station_name)%>%
  arrange(desc(n))%>%
  filter(member_casual=='member')
head(popular_end_member)


```




```{r Analyzing Data}
#1.Number of Rides per Day
data_cleaned %>%
  group_by(member_casual,day_of_week) %>%
  summarize(number_of_rides = n()) %>%
  ggplot(aes(x=day_of_week,y=number_of_rides, fill=member_casual))+
  geom_col(position = "dodge2")+
  labs(title = "Number of Rides per Day", x="Day",y="Total Rides")+
  scale_x_discrete(limits=c("Monday", "Tuesday", "Wednesday", "Thursday", "Friday", "Saturday","Sunday"))+
  theme(axis.text.x=element_text(angle=60, hjust=1))

#2.Number of Rides per Month
data_cleaned %>%
  group_by(member_casual, month) %>%
  summarise(number_of_rides=n())%>%
  ggplot (aes(x=month,y=number_of_rides, fill=member_casual))+geom_col(position = "dodge2")+
  labs(title = "Number of Rides per Month", x="Month",y="Total Rides")+
  theme(axis.text.x=element_text(angle=60, hjust=1))+
  scale_x_discrete(limits=c("January", "February", "March", "April", "May", "June", "July", "August", "September", "October", "November", "December"))

#3.Average Road Trip per Day
data_cleaned %>%
  group_by(member_casual,day_of_week) %>%
  summarize(average_trip_time=mean(trip_time)) %>%
  ggplot(aes(x=day_of_week,y=average_trip_time, fill=member_casual))+
  geom_col(position = "dodge2")+
  labs(title = "Average Road Trip per Day", x="Day",y="Total Rides")+
  scale_x_discrete(limits=c("Monday", "Tuesday", "Wednesday", "Thursday", "Friday", "Saturday","Sunday"))+
  theme(axis.text.x=element_text(angle=60, hjust=1))

#4.Average Road Trip per per Month
data_cleaned %>%
  group_by(member_casual, month) %>%
  summarise(average_trip_time=mean(trip_time))%>%
  ggplot (aes(x=month,y=average_trip_time, fill=member_casual))+geom_col(position = "dodge2")+
  labs(title = "Average Road Trip per per Month", x="Month",y="Total Rides")+
  theme(axis.text.x=element_text(angle=60, hjust=1))+
  scale_x_discrete(limits=c("January", "February", "March", "April", "May", "June", "July", "August", "September", "October", "November", "December"))

#5.Number of Rides Each Ride_type
data_cleaned %>%
  group_by(member_casual,rideable_type) %>%
  summarise(number_of_rides=n())%>%
  ggplot(aes(x=rideable_type,y=number_of_rides,fill=member_casual))+geom_col()+
  labs(title = "Number of Rides Each Ride_type", x="Ride_type",y="Total Rides")+
  theme(axis.text.x=element_text(angle=30, hjust=1))+
  facet_wrap(~member_casual)

#6.Average Road Trip Each Ride_type
data_cleaned %>%
  group_by(member_casual,rideable_type) %>%
  summarise(average_trip_time=mean(trip_time))%>%
  ggplot(aes(x=rideable_type,y=average_trip_time,fill=member_casual))+geom_col()+
  labs(title = "Average Road Trip Each Ride_type", x="Ride_type",y="Total Rides")+
  theme(axis.text.x=element_text(angle=30, hjust=1))+
  facet_wrap(~member_casual)

```


